---
title: "R Notebook"
output: html_notebook
---


##Importing tables
```{r echo = FALSE}
library(dplyr)
library(readxl)

setwd("/Users/karlavasco/Documents/")

#read in otu table
otu_table = read_excel("metabolites_classification_mastitis.xlsx", sheet="matrix")

library(tibble)
otu_table <- otu_table %>% remove_rownames %>% column_to_rownames(var="OTU")
otu_table=as.matrix(otu_table)
otu_table <- otu_table*10^9
mode(otu_table) <- "integer"

#read in taxonomy
taxonomy = read_excel("metabolites_classification_mastitis.xlsx", sheet="taxonomy")
taxonomy <- taxonomy %>% remove_rownames %>% column_to_rownames(var="OTU")
taxonomy=as.matrix(taxonomy)

#read in metadata
metadata <- read_excel("metabolites_classification_mastitis.xlsx", sheet = "metadata")
metadata <- metadata %>% remove_rownames %>% column_to_rownames(var="sample_ID")

library("phyloseq")

#import as phyloseq objects
OTU = otu_table(otu_table,taxa_are_rows=TRUE)
TAX = tax_table(taxonomy)
META = sample_data(metadata)

#Final phyloseq output is named as physeq
physeq_metabolome=phyloseq(OTU,TAX,META)
```

# Aglomerating pathways at the Subclass level
```{r}
physeq_subclass <- tax_glom(physeq_metabolome, taxrank = "Class")
```

# Converting phyloseq object to dataframe
```{r}
physeq_subclass_df <- psmelt(physeq_subclass)
```
# Summary table
```{r}
mean_metabolites <- physeq_metabolome %>% psmelt() %>% 
  group_by(Treatment, Time_tx, Class, Name, feature) %>% 
  summarise(Average = mean(Abundance)/10^9)
```
```{r}
write.csv(mean_metabolites, "/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables/mean_metabolites.csv", row.names = F)
```

# converting dataframe to matrix 
```{r}
library(tibble)
path_subclass_mx <- physeq_subclass_df %>% select(Sample, Class, Abundance) %>% pivot_wider(names_from = Class, values_from = Abundance) %>% column_to_rownames("Sample") %>% as.matrix()
```

# Annotations for Heatmap
```{r}
library(dplyr)
annotation_samples <- physeq_subclass_df %>% 
  dplyr::select(Sample, Treatment, Time_tx) %>% distinct() %>%    
  remove_rownames %>% column_to_rownames(var="Sample")


anno_color <- list(Treatment = c(Control = "#3B4992FF", Antibiotic = "#EE0000FF"),
                   Time_tx = c(`Day -1` = "#FFD700", `Week 1`= "#008000", 
                                `Week 9`="#800080"))
```

```{r fig.height=10}
library(pheatmap)
library(viridis) #color pallet, it's optional
library(RColorBrewer)

#HEATMAP RA USING ln color pallet RBrewer
heatmap <- pheatmap(
  mat               = t(log(path_subclass_mx+1)),
  border_color      = NA,
  show_colnames     = F,
  show_rownames     = T,
  angle_col = 90,
#  drop_levels       = TRUE,
 # fontsize_col = 4,
 fontsize_row = 5,
#  fontsize          = 14,
 # color             = brewer.pal(9,"RdYlBu"),
#color = inferno(100),  
#number_color = NA,
 annotation_col = annotation_samples,
  annotation_colors = anno_color,
  annotation_names_col = F,
  annotation_names_row = F,
  cluster_cols = T,
  cluster_rows = T,
 clustering_method = "ward.D2",
  gaps_row = FALSE,
)
heatmap
```
```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=heatmap, "heatmap_metabolites_mastitis.png", width = 10, height = 10)

```

# Differential abundance analyses

Aggregate phyloseq objects by time points
```{r}
physeq_day1_week1 <- subset_samples(physeq_subclass, Time_tx%in%c("Day -1","Week 1"))
physeq_day1_week9 <- subset_samples(physeq_subclass, Time_tx%in%c("Day -1","Week 9"))
physeq_week9_1 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 1","Week 9"))
```

Aggregate phyloseq objects by time points
```{r}
physeq_day1 <- subset_samples(physeq_subclass, Time_tx%in%"Day -1")
physeq_week1 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 1"))
physeq_week9 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 9"))
```

Day 0 (other time points were done in the same way)
```{r}
physeq <- physeq_day1
```


MaAsLin2
```{r results = FALSE}
library(Maaslin2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables/metabolites")
# Formating abundance table for MaAsLin2 
table <- merge(physeq@tax_table,data.frame(otu_table(physeq)), by = 0) %>% remove_rownames %>% column_to_rownames(var="Row.names")

fit_data = Maaslin2(
  input_data = merge(physeq@tax_table,data.frame(otu_table(physeq)), by = 0) %>% 
    remove_rownames %>% column_to_rownames(var="Row.names")  %>% select(5:44),
  input_metadata = metadata,
  output = 'maaslin2_day1_treatment',
  fixed_effects = c('Time_tx'),
#  random_effects = c("temperature_Celsius"),
  min_prevalence = 0.01,
  min_abundance =  0.0,
  standardize = F
)
```

# Aglomerating pathways at the Subclass level
```{r}
physeq_name <- tax_glom(physeq_metabolome, taxrank = "Name")
```

# Converting phyloseq object to dataframe
```{r}
physeq_df <- psmelt(physeq_name)
```

# Random Forest
```{r}
library(microbial)
rf_time <- biomarker(physeq=physeq_name, group = "Time_tx", normalize = F)
```

```{r}
rf_time %>% head(n=25) %>% ggbarplot(y="Value",x="Name"
                                     #, fill = "Class", palette = "simpsons"
                                     ) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
library(microbial)
rf_time <- biomarker(physeq=physeq_metabolome, group = "Time_tx", normalize = F)
```

```{r}
rf_time %>% head(n=25) %>% ggbarplot(y="Value",x="Name") + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```


# Aglomerating pathways at the Subclass level
```{r}
physeq_name <- tax_glom(physeq_metabolome, taxrank = "Name")
```

# Converting phyloseq object to dataframe
```{r}
physeq_df <- psmelt(physeq_name)
```

```{r fig.height=8, fig.width=8}
top25 <- rf_time %>% head(n=25)
mean <- physeq_df %>% group_by(Name) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Name %in% top25$Name) %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% group_by(Time_tx, Name) %>% summarise(top = mean(FC)) %>% filter(Time_tx=="Day -1") %>% arrange(top)
top_25_plot <- physeq_df %>% filter(Name %in% top25$Name) %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Name",y="FC", fill="Time_tx", palette = c("#FFD700", "#008000", "#800080"),
            add = c("mean_ci"), position = position_dodge(width = 0.7), orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            order = order$Name)
top_25_plot
```


```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=top_25_plot, "top25_rf_metabolites_mastitis.png", width = 10, height = 8)

```

```{r fig.height=10, fig.width=6}
top50 <- rf_time %>% head(n=50)
mean <- physeq_df %>% group_by(Name) %>% summarise(Average = mean(Abundance))

top_50_plot <- physeq_df %>% filter(Name %in% top50$Name) %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Name",y="FC", fill="Time_tx", add = c("mean_se"), position = position_dodge2(), orientation="horiz", facet.by = "Time_tx")
top_50_plot
```
# Converting phyloseq object to dataframe
```{r}
physeq_df <- psmelt(physeq_metabolome)
```

# converting dataframe to matrix 
```{r}
library(tibble)
mean <- physeq_df %>% group_by(feature) %>% summarise(Average = mean(Abundance))

metabolites_df <- physeq_df %>%  merge(mean, by="feature") %>% mutate(FC=Abundance/Average) %>% 
  select(Sample, feature, FC) %>% filter(feature %in% top50$feature)

metabolites_mx <- spread(metabolites_df, key=feature, value=FC,fill = 0) %>% 
  select(-Glaucine, -`histamine-C15:0`,-`L-TYROSINE - 40.0 eV`) %>% 
  column_to_rownames("Sample") %>% as.matrix() 
```

# Annotations for Heatmap
```{r}
library(dplyr)
annotation_samples <- physeq_df %>% 
  dplyr::select(Sample, Treatment, Time_tx) %>% distinct() %>%    
  remove_rownames %>% column_to_rownames(var="Sample")


anno_color <- list(Treatment = c(Control = "#3B4992FF", Antibiotic = "#EE0000FF"),
                   Time_tx = c(`Day -1` = "#FFD700", `Week 1`= "#008000", 
                                `Week 9`="#800080"))
```

```{r fig.height=10}
library(pheatmap)
library(viridis) #color pallet, it's optional
library(RColorBrewer)

#HEATMAP RA USING ln color pallet RBrewer
heatmap <- pheatmap(
  mat               = t(log10(metabolites_mx+1)),
  border_color      = NA,
  show_colnames     = F,
  show_rownames     = T,
  angle_col = 90,
#  drop_levels       = TRUE,
 # fontsize_col = 4,
 fontsize_row = 5,
#  fontsize          = 14,
 # color             = brewer.pal(9,"RdYlBu"),
#color = inferno(100),  
#number_color = NA,
 annotation_col = annotation_samples,
  annotation_colors = anno_color,
  annotation_names_col = F,
  annotation_names_row = F,
  cluster_cols = T,
  cluster_rows = T,
 clustering_method = "ward.D2",
  gaps_row = FALSE,
)
heatmap
```
```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=heatmap, "heatmap_rf_metabolites_mastitis.pdf", width = 10, height = 8)

```

# Converting phyloseq object to dataframe
```{r}
physeq_df <- psmelt(physeq_metabolome)
```

# Treatment plot
```{r}
physeq_day1 <- subset_samples(physeq_metabolome, Time_tx%in%c("Day -1"))
physeq_week1 <- subset_samples(physeq_metabolome, Time_tx%in%c("Week 1"))
physeq_week9 <- subset_samples(physeq_metabolome, Time_tx%in%c("Week 9"))
```

```{r}
library(microbial)
rf_treatment <- biomarker(physeq=physeq_week9, group = "Treatment", normalize = F)
rf_treatment
```
```{r fig.height=10}
rf_treatment %>% head(n=25) %>% ggbarplot(y="Value",x="Name"
                                     #, fill = "Class", palette = "simpsons"
                                     ) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Week 1") %>% group_by(Name) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Name %in% top25$Name, Time_tx=="Week 1") %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Name) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Control") %>% arrange(top)

top_25_w1_plot <- physeq_df %>% 
  filter(Name %in% top25$Name, Time_tx=="Week 1") %>% 
  merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Name", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Name
            )
top_25_w1_plot
```

```{r}
physeq_df %>% 
  filter(Name %in% top25$Name, Time_tx=="Week 9") %>% 
  group_by(Name) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "less") %>% 
  add_significance()
```

```{r}
physeq_df %>% 
  filter(Name %in% top25$Name, Time_tx=="Week 9") %>% 
  group_by(Name) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "greater") %>% 
  add_significance()
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Week 9") %>% group_by(Name) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Name %in% top25$Name, Time_tx=="Week 9") %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Name) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Control") %>% arrange(top)

top_25_w9_plot <- physeq_df %>% 
  filter(Name %in% top25$Name, Time_tx=="Week 9") %>% 
  merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Name", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Name
            )
top_25_w9_plot
```


```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Day -1") %>% group_by(Name) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Name %in% top25$Name, Time_tx=="Day -1") %>% merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Name) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Antibiotic") %>% arrange(top)

top_25_d1_plot <- physeq_df %>% 
  filter(Name %in% top25$Name, Time_tx=="Day -1") %>% 
  merge(mean, by="Name") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Name", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Name
            )
top_25_d1_plot
```

```{r fig.width=20}
metabolite_diff_plots <- ggarrange(top_25_w1_plot, top_25_w9_plot, labels = c("A","B"), nrow=1, common.legend = T)
metabolite_diff_plots
```

```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=metabolite_diff_plots, "top25_rf_metabolites_mastitis.png", width = 10, height = 6)
```
