---
title: "R Notebook"
output: html_notebook
---

```{r}
library(readxl)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables")
path <- read_excel("metabolic_pathways_mastitis_2023.xlsx", sheet = "taxonomy_path")
path_ig <- read_excel("metabolic_pathways_mastitis_2023.xlsx", sheet = "taxonomy_ig")

```

```{r}
pathways <- merge(path, path_ig, all.x=T, by=c("Pathway", "Description"))
```

```{r}
write.csv(pathways, row.names = F, "/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables/pathways_tax.csv")
```

##Importing tables
```{r echo = FALSE}
library(dplyr)
library(readxl)

setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables")

#read in otu table
otu_table = read_excel("metabolic_pathways_mastitis_2023.xlsx", sheet="humann_pathabundance_relab")

library(tibble)
otu_table <- otu_table %>% remove_rownames %>% column_to_rownames(var="OTU_ID")
otu_table=as.matrix(otu_table)
otu_table <- otu_table
#mode(otu_table) <- "integer"

#read in taxonomy
taxonomy = read_excel("metabolic_pathways_mastitis_2023.xlsx", sheet="taxonomy_pathways")
taxonomy <- taxonomy %>% remove_rownames %>% column_to_rownames(var="OTU_ID")
taxonomy=as.matrix(taxonomy)

#read in metadata
metadata <- read_excel("metabolic_pathways_mastitis_2023.xlsx", sheet = "metadata")
metadata <- metadata %>% remove_rownames %>% column_to_rownames(var="sequence_id")

library("phyloseq")

#import as phyloseq objects
OTU = otu_table(otu_table,taxa_are_rows=TRUE)
TAX = tax_table(taxonomy)
META = sample_data(metadata)

#Final phyloseq output is named as physeq
physeq_pathways=phyloseq(OTU,TAX,META)
```

# Aglomerating pathways at the Subclass level
```{r}
physeq_subclass <- tax_glom(physeq_pathways, taxrank = "Subclass")
```

# Converting phyloseq object to dataframe
```{r}
physeq_subclass_df <- psmelt(physeq_subclass)
```
```{r}
mean_pathways <- physeq_pathways %>% psmelt() %>% 
  group_by(Treatment, Time_tx, Class, Subclass, Pathway, Description) %>% 
  summarise(Average = mean(Abundance))
```

```{r}
write.csv(mean_pathways, "/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables/mean_pathways.csv", row.names = F)
```

# converting dataframe to matrix 
```{r}
library(tibble)
path_subclass_mx <- physeq_subclass_df %>% select(sample_ID, Subclass, Abundance) %>% pivot_wider(names_from = Subclass, values_from = Abundance) %>% select(-UNMAPPED,-UNINTEGRATED) %>% column_to_rownames("sample_ID") %>% as.matrix()
```

# Annotations for Heatmap
```{r}
library(dplyr)
annotation_samples <- physeq_subclass_df %>% 
  dplyr::select(sample_ID, Treatment, Time_tx) %>% distinct() %>%    
  remove_rownames %>% column_to_rownames(var="sample_ID")


anno_color <- list(Treatment = c(Control = "#3B4992FF", Antibiotic = "#EE0000FF"),
                   Time_tx = c(`Day -1` = "#FFD700", `Week 1`= "#008000", 
                               `Week 5` = "#8a9197ff", `Week 9`="#800080"))
```

```{r}
library(pheatmap)
library(viridis) #color pallet, it's optional
library(RColorBrewer)

#HEATMAP RA USING ln color pallet RBrewer
heatmap <- pheatmap(
  mat               = t(log10(path_subclass_mx+1)),
  border_color      = NA,
  show_colnames     = F,
  show_rownames     = T,
  angle_col = 90,
#  drop_levels       = TRUE,
 # fontsize_col = 4,
 fontsize_row = 5,
#  fontsize          = 14,
 # color             = brewer.pal(9,"RdYlBu"),
#color = inferno(100),  
#number_color = NA,
 annotation_col = annotation_samples,
  annotation_colors = anno_color,
  annotation_names_col = F,
  annotation_names_row = F,
  cluster_cols = T,
  cluster_rows = T,
 clustering_method = "ward.D2",
  gaps_row = FALSE,
)
heatmap
```
```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=heatmap, "heatmap_pathways_mastitis.png", width = 10, height = 5)

```

# Differential abundance analyses

Aggregate phyloseq objects by time points
```{r}
physeq_day1_week1 <- subset_samples(physeq_subclass, Time_tx%in%c("Day -1","Week 1"))
physeq_day1_week9 <- subset_samples(physeq_subclass, Time_tx%in%c("Day -1","Week 9"))
physeq_day1_week5 <- subset_samples(physeq_subclass, Time_tx%in%c("Day -1","Week 5"))
physeq_week9_5 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 9","Week 1"))
physeq_week9_1 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 9","Week 5"))
```

Aggregate phyloseq objects by time points
```{r}
physeq_day1 <- subset_samples(physeq_subclass, Time_tx%in%"Day -1")
physeq_week1 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 1"))
physeq_week5 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 5"))
physeq_week9 <- subset_samples(physeq_subclass, Time_tx%in%c("Week 9"))
```

Day 0 (other time points were done in the same way)
```{r}
physeq <- physeq_week5
```


MaAsLin2
```{r results = FALSE}
library(Maaslin2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/tables/pathways")
# Formating abundance table for MaAsLin2 
table <- merge(physeq@tax_table,data.frame(otu_table(physeq)), by = 0) %>% remove_rownames %>% column_to_rownames(var="Row.names")

fit_data = Maaslin2(
  input_data = merge(physeq@tax_table,data.frame(otu_table(physeq)), by = 0) %>% 
    remove_rownames %>% column_to_rownames(var="Row.names")  %>% select(14:53),
  input_metadata = metadata,
  output = 'maaslin2_day1_treatment',
  fixed_effects = c('Treatment'),
#  random_effects = c("temperature_Celsius"),
  min_prevalence = 0.01,
  min_abundance =  0.0,
  standardize = F
)
```

# Aglomerating pathways at the Subclass level
```{r}
physeq_path <- tax_glom(physeq_pathways, taxrank = "Description")
```

```{r}
library(microbial)
rf_time <- biomarker(physeq=physeq_path, group = "Time_tx", normalize = F)
rf_time
```


```{r fig.height=20}
rf_time %>% head(n=25) %>% ggbarplot(y="Value",x="Description"
                                     #, fill = "Class", palette = "simpsons"
                                     ) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# Converting phyloseq object to dataframe
```{r}
physeq_df <- psmelt(physeq_path)
```

```{r}
library(data.table)
physeq_df %>% 
  filter(Time_tx %in% c("Week 5","Week 9")) %>% 
  group_by(Description, Time_tx,Treatment) %>% 
 summarise(count = sum(Abundance > 0)) %>% 
  filter(Description %like% "L-ornithine biosynthesis II")
```


```{r fig.height=12, fig.width=8}
top25 <- rf_time %>% head(n=25)
mean <- physeq_df %>% group_by(Description) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Description %in% top25$Description) %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% group_by(Time_tx, Description) %>% summarise(top = mean(FC)) %>% filter(Time_tx=="Day -1") %>% arrange(top)

top_25_plot <- physeq_df %>% filter(Description %in% top25$Description) %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Description",y="FC", fill="Time_tx", palette = c("#FFD700", "#008000", "#0073C2ff", "#800080"),
            add = c("mean_ci"), position = position_dodge(width = 0.7), orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            #facet.by = "Class",
            order = order$Description)
top_25_plot
```

```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=top_25_plot, "top25_rf_pathways_mastitis.png", width = 10, height = 8)
```

# Treatment plot
```{r}
physeq_day1 <- subset_samples(physeq_path, Time_tx%in%c("Day -1"))
physeq_week1 <- subset_samples(physeq_path, Time_tx%in%c("Week 1"))
physeq_week5 <- subset_samples(physeq_path, Time_tx%in%c("Week 5"))
physeq_week9 <- subset_samples(physeq_path, Time_tx%in%c("Week 9"))
```

```{r}
library(microbial)
rf_treatment <- biomarker(physeq=physeq_week1, group = "Treatment", normalize = F)
rf_treatment
```
```{r fig.height=10}
rf_treatment %>% head(n=25) %>% ggbarplot(y="Value",x="Description"
                                     #, fill = "Class", palette = "simpsons"
                                     ) + theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Week 5") %>% group_by(Description) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Description %in% top25$Description, Time_tx=="Week 5") %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Description) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Control") %>% arrange(top)

top_25_w5_plot <- physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 5") %>% 
  merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Description", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Description
            )
top_25_w5_plot
```

```{r}
physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 1") %>% 
  group_by(Description) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "less") %>% 
  add_significance()

physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 1") %>% 
  group_by(Description) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "greater") %>% 
  add_significance()
```

```{r}
physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 5") %>% 
  group_by(Description) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "greater") %>% 
  add_significance()
```
```{r}
physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 9") %>% 
  group_by(Description) %>% 
  wilcox_test(Abundance ~ Treatment, paired = F, alternative = "greater") %>% 
  add_significance()
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Week 9") %>% group_by(Description) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Description %in% top25$Description, Time_tx=="Week 9") %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Description) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Control") %>% arrange(top)

top_25_w9_plot <- physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 9") %>% 
  merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Description", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Description
            )
top_25_w9_plot
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Week 1") %>% group_by(Description) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Description %in% top25$Description, Time_tx=="Week 1") %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Description) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Antibiotic") %>% arrange(top)

top_25_w1_plot <- physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Week 1") %>% 
  merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Description", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            facet.by = "Time_tx",
            order = order$Description
            )
top_25_w1_plot
```

```{r fig.height=12, fig.width=8}
top25 <- rf_treatment %>% head(n=25)
mean <- physeq_df %>% filter(Time_tx=="Day -1") %>% group_by(Description) %>% summarise(Average = mean(Abundance))

order <- physeq_df %>% filter(Description %in% top25$Description, Time_tx=="Day -1") %>% merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% group_by(Treatment, Description) %>% summarise(top = mean(FC)) %>% filter(Treatment=="Antibiotic") %>% arrange(top)

top_25_d1_plot <- physeq_df %>% 
  filter(Description %in% top25$Description, Time_tx=="Day -1") %>% 
  merge(mean, by="Description") %>% mutate(FC=Abundance/Average) %>% 
  ggbarplot(x="Description", y="FC", fill="Treatment", palette = "npg",
            add = c("mean_ci"), position = position_dodge(width = 0.7), 
            orientation="horiz", xlab = "", ylab = "Fold change (mean)",
            #facet.by = "Class",
            order = order$Description
            )
top_25_d1_plot
```

```{r fig.width=20}
w1_w5_w9_diff_plots <- ggarrange(top_25_w1_plot, top_25_w5_plot, top_25_w9_plot, labels = c("A","B","C"), nrow=1, common.legend = T)
w1_w5_w9_diff_plots
```

```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=w1_w5_w9_diff_plots, "top25_rf_metabolicpaths_mastitis_final.png", width = 32, height = 8)
```
# Pathways differential
```{r fig.width=20}
path_diff_plots <- ggarrange(top_25_w1_plot,top_25_w5_plot, top_25_w9_plot, labels = c("A","B","C"), nrow=1, common.legend = T)
path_diff_plots
```

```{r}
library(ggplot2)
setwd("/Users/karlavasco/Library/CloudStorage/OneDrive-MichiganStateUniversity/Manning_lab/Metabolomics_mastitis/figures")
ggsave(plot=path_diff_plots, "top_rf_metabolicpaths_mastitis.png", width = 24, height = 8)
```